---
title: "R Notebook"
output: html_notebook
---

```{r}
library(MOFA2)
library(MOFAdata)
library(data.table)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(data.table)
library(Hmisc)
library(matrixStats)

source("utils.R")


# Meta Data
wes_clinical <- read_tsv('/oak/stanford/groups/ccurtis2/users/azizk/projects/FAP/data/calls_for_MOFA/wes_clinical.tsv') # META DATA
wgs_clinical <- read_tsv('/oak/stanford/groups/ccurtis2/users/azizk/projects/FAP/data/calls_for_MOFA/wgs_clinical.tsv') # META DATA

exonic_mut <- read_tsv("HTAN_exonic_mutation_burden.tsv")
FAP_SV_burden <- read_tsv("HTAN_FAP_SV_burden.tsv")
percent_genome_altered <- read_tsv("HTAN_percent_genome_altered.tsv")


met <- make_meta(wes_clinical, wgs_clinical, exonic_mut, FAP_SV_burden, percent_genome_altered)

#head(met)
met_stage <- met %>% dplyr::select(sample, stage)
met_stage

rna_org <- read_tsv("vst_batch_covars.stage.tsv") 
rna_df <- make_rna_seq_df(rna_org)
dt <- as.data.table(rna_df)
dt
rna_dt <- dt[, mean(value), by = list(sample, feature, view)]
colnames(rna_dt) <- c("sample", "feature", "view", "value")
rna_dt_fin <- rna_dt %>% dplyr::select(sample, feature, value, view)

rna_df <- rna_dt_fin


rna_df_DE <- n_most_var_df(rna_df, 466)




wes_gene_cna <- read_tsv('/oak/stanford/groups/ccurtis2/users/azizk/projects/FAP/data/calls_for_MOFA/Filtered_wes_gene_level_cna_facets.tsv') # CNA
wgs_gene_cna <- read_tsv('/oak/stanford/groups/ccurtis2/users/azizk/projects/FAP/data/calls_for_MOFA/Filtered_wgs_gene_level_cna_facets.tsv') # CNA



CNA_original <- merge_wes_wgs_cna(wes_gene_cna, wgs_gene_cna)
CNA_original <- as.data.frame(CNA_original)


CNA <- n_most_var_df(CNA_original, 466)


wes_snv_maf <- read_tsv('/oak/stanford/groups/ccurtis2/users/azizk/projects/FAP/data/calls_for_MOFA/Filtered_wes_snv_maf.tsv') # SNV
wgs_snv_maf <- read_tsv('/oak/stanford/groups/ccurtis2/users/azizk/projects/FAP/data/calls_for_MOFA/Filtered_wgs_snv_maf.tsv') # SNV

genes <- c("APC","AMER1", "TCF7L2", "CTNNB1","KRAS", "BRAF", "NRAS","TP53","FBXW7","PIK3CA", "PTEN","SMAD4", "SMAD2", "ACVR2A","GNAS","ARID1A", "SOX9", "TGIF1", "PCBP1", "ZFP36L2") # PanCancer Genes


snv <- make_snv(wes_snv_maf, wgs_snv_maf)
SNV <- snv_thresh(snv, 0.25)
SNV <- SNV[!SNV$feature == 'Unknown', ]


SNV$view <- "SNV"
SNV <- SNV %>% dplyr::select(sample, feature, value, view)
CNA$view <- "CNA"
rna_df_DE$view <- "RNA"

rna_df_DE

mof_data <- rbind(SNV, CNA, rna_df_DE)

SNV_t <- merge(SNV, met_stage)
SNV_t <- SNV_t %>% dplyr::select(sample, feature, value, stage, view)
colnames(SNV_t) <- c("sample", "feature", "value", "group", "view")

mof_data_grouped <- merge(mof_data, met_stage)
mof_data_grouped
mof_data_grouped <- mof_data_grouped %>% dplyr::select(sample, feature, value, stage, view)
colnames(mof_data_grouped) <- c("sample", "feature", "value", "group", "view")

mof_data_grouped
MOFAobject <- create_mofa(mof_data_grouped)
plot_data_overview(MOFAobject)

```




```{r}
data_opts <- get_default_data_options(MOFAobject)
data_opts

model_opts <- get_default_model_options(MOFAobject)
model_opts$num_factors <- 15

model_opts

train_opts <- get_default_training_options(MOFAobject)
train_opts$convergence_mode <- "slow"
train_opts$seed <- 42

train_opts


MOFAobject <- prepare_mofa(MOFAobject,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)

#MOFAobject <- run_mofa(MOFAobject, outfile="MOFA2_Group_Stage_RNA.hdf5")
#saveRDS(MOFAobject,"MOFA2_Group_Stage_RNA.hdf5")

MOFAobject <- readRDS("MOFA2_Group_Stage.hdf5")
plot_data_overview(MOFAobject)
```

#  variance decomposition by factor, total variance explained across views   

```{r}
plot_variance_explained(MOFAobject, x="group", y="factor")
plot_variance_explained(
  MOFAobject, 
  x = "group", 
  y = "factor", 
  plot_total = T
)[[2]]
```

```{r}
head(met) 
colnames(met)[6] <- "group"
samples_metadata(MOFAobject) <- met
met
```

```{r}
correlate_factors_with_covariates(MOFAobject, 
  covariates = c("total_perMB", "log2_svc_burden", "pct_change", "Patient", "Location"), 
  plot="log_pval"
)
```




```{r}
plot_factor(MOFAobject, 
  factor = 1,
  color_by = "Patient"  # lineage is a column in MOFAobject@samples.metadata
) #+ scale_fill_manual(values=colors)
```

```{r}
plot_factor(MOFAobject,
  factor = 1,
  color_by = "Patient", 
  scale = TRUE, 
  add_violin = TRUE, color_violin = TRUE, 
  dodge = TRUE, dot_size = 1, legend = TRUE
) #+ scale_color_manual(values=colors) + scale_fill_manual(values=colors)
```
# Factor 1 analysis via SNV view
### Factor 1 sample values vs 4 top 6 weighted genes : stage = mucosa
```{r}
p <- plot_data_scatter(MOFAobject, 
  view = "SNV", 
  factor = 1, 
  features = 6,         # Number of features to show
  sign = "positive",     # select top 6 features with positive weights
  color_by = "Patient",  # color cells by lineage
  add_lm = T,          # add linear regression estimates
  lm_per_group = F, 
  dot_size = 2,
  group = "Mucosa"
)


print(p)
```
### Factor 1 sample values vs 4 top 6 weighted genes : stage = benign
```{r}
p <- plot_data_scatter(MOFAobject, 
  view = "SNV", 
  factor = 1, 
  features = 6,         # Number of features to show
  sign = "positive",     # select top 6 features with positive weights
  color_by = "Patient",  # color cells by lineage
  add_lm = T,          # add linear regression estimates
  lm_per_group = F, 
  dot_size = 2,
  group = "Benign"
)


print(p)
```
### Factor 1 sample values vs 4 top 6 weighted genes : stage = dysplasia
```{r}
p <- plot_data_scatter(MOFAobject, 
  view = "SNV", 
  factor = 1, 
  features = 6,         # Number of features to show
  sign = "positive",     # select top 6 features with positive weights
  color_by = "Patient",  # color cells by lineage
  add_lm = T,          # add linear regression estimates
  lm_per_group = F, 
  dot_size = 2,
  group = "Dysplasia"
)


print(p)
```
### Factor 1 sample values vs 4 top 6 weighted genes : stage = adenocarcinoma

```{r}
p <- plot_data_scatter(MOFAobject, 
  view = "SNV", 
  factor = 1, 
  features = 6,         # Number of features to show
  sign = "positive",     # select top 6 features with positive weights
  color_by = "Patient",  # color cells by lineage
  add_lm = T,          # add linear regression estimates
  lm_per_group = F, 
  dot_size = 2,
  group = "AdCa"
)


print(p)
```
# Factor 1 analysis - CNA view
### Factor 1 sample values vs 4 top 6 weighted genes : stage = Mucosa
```{r}
p <- plot_data_scatter(MOFAobject, 
view = "CNA", 
  factor = 1, 
  features = 6,         # Number of features to show
  sign = "positive",     # select top 6 features with positive weights
  color_by = "Patient",  # color cells by lineage
  add_lm = T,          # add linear regression estimates
  lm_per_group = F, 
  dot_size = 2,
  group = "Mucosa"
)


print(p)
```
### Factor 1 sample values vs 4 top 6 weighted genes : stage = Benign
```{r}
p <- plot_data_scatter(MOFAobject, 
view = "CNA", 
  factor = 1, 
  features = 6,         # Number of features to show
  sign = "positive",     # select top 6 features with positive weights
  color_by = "Patient",  # color cells by lineage
  add_lm = T,          # add linear regression estimates
  lm_per_group = F, 
  dot_size = 2,
  group = "Benign"
)


print(p)
```
### Factor 1 sample values vs 4 top 6 weighted genes : stage = Dysplasia

```{r}
p <- plot_data_scatter(MOFAobject, 
view = "CNA", 
  factor = 1, 
  features = 6,         # Number of features to show
  sign = "positive",     # select top 6 features with positive weights
  color_by = "Patient",  # color cells by lineage
  add_lm = T,          # add linear regression estimates
  lm_per_group = F, 
  dot_size = 2,
  group = "Dysplasia"
)


print(p)
```
### Factor 1 sample values vs 4 top 6 weighted genes : stage = AdCa
```{r}
p <- plot_data_scatter(MOFAobject, 
view = "CNA", 
  factor = 1, 
  features = 6,         # Number of features to show
  sign = "positive",     # select top 6 features with positive weights
  color_by = "Patient",  # color cells by lineage
  add_lm = T,          # add linear regression estimates
  lm_per_group = F, 
  dot_size = 2,
  group = "AdCa"
)


print(p)
```
# Factor 2 analysis - SNV view
```{r}
plot_factor(MOFAobject,
  factor = 2,
  color_by = "Patient", 
  scale = TRUE, 
  add_violin = TRUE, color_violin = TRUE, 
  dodge = TRUE, dot_size = 1, legend = TRUE
) #+ scale_color_manual(values=colors) + scale_fill_manual(values=colors)
```


### Factor 2 sample values vs 4 top 6 weighted genes : stage = Mucosa
```{r}
p <- plot_data_scatter(MOFAobject, 
view = "SNV", 
  factor = 2, 
  features = 6,         # Number of features to show
  sign = "positive",     # select top 6 features with positive weights
  color_by = "Patient",  # color cells by lineage
  add_lm = T,          # add linear regression estimates
  lm_per_group = F, 
  dot_size = 2,
  group = "Mucosa"
)


print(p)
```
### Factor 2 sample values vs 4 top 6 weighted genes : stage = Benign
```{r}
p <- plot_data_scatter(MOFAobject, 
view = "SNV", 
  factor = 2, 
  features = 6,         # Number of features to show
  sign = "positive",     # select top 6 features with positive weights
  color_by = "Patient",  # color cells by lineage
  add_lm = T,          # add linear regression estimates
  lm_per_group = F, 
  dot_size = 2,
  group = "Benign"
)


print(p)
```

### Factor 2 sample values vs 4 top 6 weighted genes : stage = Mucosa

```{r}
p <- plot_data_scatter(MOFAobject, 
view = "SNV", 
  factor = 2, 
  features = 6,         # Number of features to show
  sign = "positive",     # select top 6 features with positive weights
  color_by = "Patient",  # color cells by lineage
  add_lm = T,          # add linear regression estimates
  lm_per_group = F, 
  dot_size = 2,
  group = "Dysplasia"
)


print(p)
```

```{r}
p <- plot_data_scatter(MOFAobject, 
view = "SNV", 
  factor = 2, 
  features = 6,         # Number of features to show
  sign = "positive",     # select top 6 features with positive weights
  color_by = "Patient",  # color cells by lineage
  add_lm = T,          # add linear regression estimates
  lm_per_group = F, 
  dot_size = 2,
  group = "AdCa"
)


print(p)
```

